---
title: "Lab 03 - Exploratory Data Analysis"
author: "RunshiYang"
output:
  tufte::tufte_html:
    css: style.css
link-citations: yes
---

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
# install.packages(c("data.table","leaflet"))
library(data.table)
library(leaflet)
library(tidyverse)
```

# Learning Goals

- Read in and get familiar with the meteorology dataset  
- Step through the EDA "checklist" presented in the class slides
- Practice making exploratory graphs

# Lab Description

We will work with the meteorological data presented in lecture. Recall the dataset consists of weather station readings in the continental US. 

**The objective of the lab is to find the weather station with the highest elevation and look at patterns in the time series of its wind speed and temperature.**

# Steps

### 1. Read in the data

First download and then read in with data.table:fread()

```{r checking-file, include=FALSE}
fn <- "https://raw.githubusercontent.com/JSC370/jsc370-2023/main/labs/lab03/met_all.gz"
if (!file.exists("met_all.gz"))
  download.file(fn, destfile = "met_all.gz")
met <- data.table::fread("met_all.gz")
```


```{r, echo=TRUE, message=FALSE, eval=FALSE}
download.file(
  "https://raw.githubusercontent.com/JSC370/jsc370-2023/main/labs/lab03/met_all.gz",
  destfile = "met_all.gz",
  method   = "curl",
  timeout  = 60
  )

met <- data.table::fread("met_all.gz")
```

### 2. Check the dimensions, headers, footers. How many columns, rows are there?

```{r}
dim(met)
head(met)
tail(met)
```
There are 2,377,343 rows and 30 columns in the `met` dataset.

### 3. Take a look at the variables.

```{r}
str(met)
```


### 4. Take a closer look at the key variables. 

```{r}
table(met$year)
table(met$day)
table(met$hour)
summary(met$temp)
summary(met$elev)
summary(met$wind.sp)
```

It looks like the elevation variable has observations with 9999.0, which is probably an indicator for missing. We should take a deeper look at the data dictionary to confirm. The wind speed variable is ok but there are a lot of missing data.

After checking the data we should make the appropriate modifications. Replace elevations with 9999 as `NA`.

```{r}
# base R
met$elev[met$elev == 9999.0] <- NA

# tidyverse
met <- met %>%
  mutate(elev = ifelse(elev == 9999, NA, elev))

# data.table
met[elev == 9999, elev := NA]

summary(met$elev)
met[elev == 4113]

```

At what elevation is the highest weather station?

- The highest weather station is the station with USAFID 720385, height 4113.

We also have the issue of the minimum temperature being -40C, so we should remove those observations. 

```{r}
table(met$temp == -40, useNA = "always")
met <- met[temp > -40]
sum(is.na(met$temp))
```


### 5. Check the data against an external data source.

We should check the suspicious temperature value (where is it located?) and validate that the range of elevations make sense (-13 m to 4113 m).

Google is your friend here.

Fix any problems that arise in your checks.

```{r}
# <place your code here>
unique(
  met[temp == max(temp), .(lat, lon, hour, elev, temp)]
)

unique(
  met[temp == min(temp), .(lat, lon, hour, elev, temp)]
)

met <- met[temp < 56]
```

- For the data point with temperature 56, it is located near the Auburn Municipal Airport, and there is only one hour of this temperature, so this is very likely to be a problematic data point.

### 6. Calculate summary statistics

Remember to keep the initial question in mind. We want to pick out the weather station with maximum elevation and examine its wind speed and temperature.

Some ideas: 
1. select the weather station with maximum elevation; 
2. look at the correlation between temperature and wind speed; and 
3. look at the correlation between temperature and wind speed with hour and day of the month.

```{r}
highest <- met[elev == max(elev, na.rm = TRUE)]
highest
unique(highest$USAFID)
cor(highest$temp, highest$wind.sp, use="complete.obs")
cor(highest[, c('temp', 'wind.sp', 'day', 'hour')], use="complete.obs")
```

### 7. Exploratory graphs

We should look at the distributions of all of the key variables (elevation, temp, wind speed) to make sure there are no remaining issues with the data.

```{r }
boxplot(met$elev, col = "pink", main = "Boxplot of elevation")
hist(met$temp, main = "Histogram of Temperature", xlab = "Temperature", ylab = "Count", col = "pink")
hist(met$wind.sp, main = "Histogram of Wind Speed", xlab = "Wind Speed", ylab = "Count", col = "pink")
```

One thing we should consider for later analyses is to log transform wind speed and elevation as the are very skewed.

Look at where the weather station with highest elevation is located (i.e. make a map!)

```{r}
leaflet(highest) %>%
  addProviderTiles('OpenStreetMap') %>%
  addCircles(lng = ~lon, lat = ~lat, color = "red", radius = 50)
```

- The station is located near the Coldrado Mines Peak and Grand Country with longtitude and latitude -105.766 and 39.8.


Look at the time series of temperature and wind speed at this location. For this we will need to create a date-time variable for the x-axis.

```{r message=FALSE}
library(lubridate)

highest$date <- ymd_h(
  paste(highest$year, highest$month, highest$day, highest$hour))

highest <- highest %>%
  mutate(date = ymd_h(paste(year, month, day, hour)))

summary(highest$date)
highest
```

With the date-time variable we can plot the time series of temperature and wind speed.

```{r}
library(ggplot2)
ggplot(data = highest, mapping = aes(x = date, y = temp)) +
  geom_line() +
  labs(title = "Temperature and Date-time", xlab  = "Date-time", y = "Temperature (deg C)")


ggplot(data = highest, mapping = aes(x = date, y = wind.sp)) +
  geom_line() +
  labs(title = "Wind Speed and Date-time", xlab  = "Date-time", y = "Wind Speed")


ggplot(highest %>% arrange(date)) +
  theme_minimal() +
  geom_line(aes(x = hour(date), y = temp,
                group = yday(date),
                colour = yday(date)),
            show.legend = FALSE) +
  scale_color_viridis_c()


ggplot(highest %>% arrange(date)) +
  theme_minimal() +
  geom_line(aes(x = hour(date), y = wind.sp,
                group = yday(date),
                colour = yday(date)),
            show.legend = FALSE) +
  scale_color_viridis_c()
            
```

Summarize any trends that you see in these time series plots.

- The temperature gradually decreases from midnight to sunrise each day, and then increases until about 14:00. After that, it starts to decrease. And I did not notice a trend in wind speed change from day to day. However, the range of daily wind speed variation became larger and larger from August 5 to August 26.
